Autonomous Oceanographic Data Collection towards Estimating Excursion Sets of Vector-valued Random Fields

### 2020-06-19 section2_old.tex
Updates from Overleaf

\subsection{Uncertainty Quantification on Excursion Sets of multivariate Gaussian Random Fields}
\label{sec:set_uq}
\input{UQonSets}

\subsection{Expected Integrated Bernoulli Variance and Excursion Measure}
\label{sec:eibv}
\input{eibv}

>>>>

\subsection{Uncertainty Quantification on Excursion Sets of multivariate Gaussian Random Fields}
\label{sec:set_uq}
We now introduce quantities that allow to quantify the uncertainty on the volume of the excursion set $\es$. Let $\mes$ be a 
(locally finite, Borel) measure  on $\domain$. We want to investigate the probability distribution 
of $\mes(\es)$ through its moments.

Centred moments may be computed using Proposition~\ref{propo1} developed in the appendix. 
In particular, it allows to write the excursion measure variance $\emv = \operatorname{Var}[\mes(\es)]$ as an integral of the excursion probability
\begin{equation*}
\begin{split}
\emv
&=\int_{\domain^2} \mathbb{P}\left(
\gp[u]\in T, \gp[v]\in T \right)
d\mes^{\otimes}(u, v)\\
&-\left( \int_{\domain} \mathbb{P}\left(\gp[u]\in T\right) d\mes(u) \right)^2,
\end{split}
\end{equation*}
which boils down in the excursion/sojourn case where $\T=(-\infty, t_1] \times
\dots \times (-\infty, t_{\no}]$ to
\begin{equation*}
\begin{split}
\emv
%\operatorname{Var}[\mes(\es)]
&=\int_{\domain^2}
\varPhi_{2\no}
\left(
(\bt, \bt); \mu((u,v)),
K((u,v),(u,v))
\right)
\
\mathrm{d}\mes^{\otimes} %\mes
%\productMeasure
(u,v)\\
&-\left( \int_{\domain} \varPhi_{\no}\left(\bt;\mu(u), K(u)\right) d\mes(u) \right)^2,
\end{split}
\end{equation*}
%
Note that like in the case of scalar-valued fields, this quantity requires to work out an integral over $\domain^2$. In 
contrast, and still like in the scalar-valued case, the Integrated Bernoulli Variance (IBV) of \cite{Bect.etal} involves 
solely an integral on $\domain$ and can be expanded in our present settings as follows
\begin{equation*}
\begin{split}
\operatorname{IBV} %(\es) %;\mes)
&=\int_{\domain}
\mathbb{P}\left(\gp[\uu]\in T\right)(1-\mathbb{P}\left(\gp[\uu]\in T\right))
d\mes(u) \\
&=\int_{\domain}
\varPhi_{\no}\left(\bt;\mu(\uu), K(\uu)\right)
-\left(\varPhi_{\no}\left(\bt;\mu(\uu), K(\uu)\right) \right)^2
\mathrm{d}\mes(u).
\end{split}
\end{equation*}
%
Next we investigate uncertainty reduction criteria related to those indicators.

\subsection{Expected Integrated Bernoulli Variance and Excursion Measure}
\label{sec:eibv}
This section is devoted to computing the expected effect of the inclusion of new observations on the $\emv$ and $\ibv$ of the excursion set $\es$. Let us consider the same setting as in Section \ref{sec:cokriging} where $n$ batches of measurements have already been performed, and let 
$\currentExp{.}$ and $\currentProba{.}$ denote conditional expectation
and probability conditional on the first $n$ batches of observations, respectively. We want to co pute the effect of the inclusion of a new set of observations at $\bm{x}_{n+1}$ on the $\emv$ and $\ibv$.

For uncertainty functional such as $\ibv$ and $\emv$, we use a similar indexing system
with respect to a sequence of batch, and further indicate when applicable the new batch
of generalized location between parentheses; e.g., $\emv_{[n+1]}(\bm{x})$ denotes the value
of $\emv$ (a random variable, seen from ``step'' $n$, i.e. when knowing $Z$ solely at
the $n$ first generalized location batches) when adding observations at $\bm{x}$ to those
already available at step $n$.

\textcolor{red}{C.T.: This doesn't work, we cannot be that short on the definition of EIBV, it is highly unclear. I suggest we go for the following notation:
$$\operatorname{IBV}(\bm{x}; \bm{y})
$$ denotes the IBV under the current law of the field, conditioned on observing $\bm{y}$ at $\bm{x}$ (generalized, possibly batch observation). Then
$$
\operatorname{EIBV}(\bm{x}):=\mathbb{E}_{\bm{Y}}\left[\textrb{IBV}(\bm{x}; \bm{Y})\right]
$$
where $\bm{Y}$ is distributed according to the current law of $Z_{\bm{x}}$.}

\textcolor{red}{
The philosophy is the following: EIBV and EEMV are functionals of a probability distribution. Then, if not mentioned explicitly, the probability distribution will be taken to be the current law of the field. (This notation is highly efficient).}

\textcolor{red}{When one want to consider several steps, one can use $\textrm{IBV}_{n}(\bm{x}; \bm{y})$ to denote the fact that the starting point (starting probability distribution) is the law of the field at the $n$-th step $\mathbb{P}_n$ to which conditioning on $\bm{y}$ at $\bm{x}$ is added.}

\textcolor{red}{For a rigorous definition
$$
\operatorname{IBV}(\bm{x}; \bm{y}):=\int_{\domain}
\mathbb{P}\left(\gp[\uu]\in T\mid Z_{\bm{x}}=\bm{y}\right)(1-\mathbb{P}\left(\gp[\uu]\in T \mid Z_{\bm{x}}=\bm{y}\right))
d\mes(u) \\
$$
where $\mathbb{P}$ denotes the current law of the field.}

\begin{propo}
\label{propo_eibv}
Let $\eibv_{[n]}(\bm{x})=\currentExp{\ibv_{[n+1]}(\bm{x})}$ (where $\bm{x}$ plays
the role of $\bm{x}_{n+1}$) be the expected effect of adding an observation a $\bm{x}$ on the IBV.
Then we have
\begin{equation}
\begin{split}
\eibv_{[n]}(\bm{x})
&=\int_{D} \varPhi_{\no}\left(\bt;~\currentMean{\uu}, \currentCov{u, u}\right) d\mes(u)\\
&-\int_{D} \varPhi_{2\no}
\left(
\left(
\begin{matrix}
\bt-\currentMean{u}\\
\bt-\currentMean{u}
\end{matrix}
\right);
\mathbf{\Sigma}_{[n]}(\uu)
\right)
d\mes(u),
\end{split}
\end{equation}
where the matrix $\mathbf{\Sigma}_{[n]}(\uu)$ is defined as
\begin{equation*}
\begin{split}
\mathbf{\Sigma}_{[n]}(\uu)&=
\left(
\begin{matrix}
\currentCov{u, u} & \currentCov{u, u}-\futureCov{u, u}\\
\currentCov{u, u}-\futureCov{u, u} & \currentCov{u, u}
\end{matrix}
%\begin{matrix}
%\currentCov{u, u} & \Delta_{[n]}(\uu)\\
%\Delta_{[n]}(\uu) & \currentCov{u, u}
%\end{matrix}
\right).\\
\end{split}
\end{equation*}
\end{propo}

As for the expected excursion measure variance, a similar result may be derived.
\begin{propo}
\label{propo_emv}

\begin{equation*}
\begin{split}
\eemv_{[n]}(\bm{x})
&=\int_{D^2} 
\varPhi_{2\no}
\left(
(\bt, \bt); \mu((u,v)), 
K((u,v),(u,v))
\right) 
\
\mathrm{d}\mes^{\otimes} %\mes 
%\productMeasure 
(u,v)\\
&-\currentExp{
    \varPhi_{\no}\left(a_1+B_1 V; C_1\right)
    \varPhi_{\no}\left(a_2+B_2 V; C_2\right)
},
\end{split}
\end{equation*}

with $V=\gp[\bm{x}_{n+1}]-\currentMean{\bm{x}_{n+1}} \sim \mathcal{N}(0_{q_{n+1}},k_{[n]}(\bm{x}_{n+1},\bm{x}_{n+1}))$ and $a_1=\bt-\currentMean{\uu}$,
$B_1=-\lambda_{[n+1,n+1]}(\uu)^T$, $a_2=\bt-\currentMean{\vv}$, $B_2=-\lambda_{[n+1,n+1]}(\vv)^T$.

This integrand finally boils down to an evaluation of the $2\no$-dimensional Gaussian CDF by
Proposition~\ref{propo3}. Details are omitted here for brevity as we did not implement further
this criterion that requires numerical integration over $\domain^2$.
\end{propo}

Note that the calculation of the multivariate Gaussian cumulative distribution
may be done effectively using code such as that of
\cite{genz2009computation}.

\textcolor{red}{Do we really want to keep this?}
A critical element in the derivation of a closed-form for the EIBV is
that the conditional mean in the cokriging equations is a linear
(affine) function of the data $\by_d$ and the covariance is not a
function of the data. This in turn means that the probabilities in Eq.
\eqref{eq:post_ep} involve inequality statements for linear
combinations of Gaussian variables. Related closed-form solutions have
been noted in similar contexts \citep{bhattacharjya2013value,
  chevalier2014fast,stroh}, but not generalized to our situation with
random sets for vector-valued GPs.

### 2020-06-19 section3.tex
Updates from Overleaf

In a sequential setting, the point of view is that $n$ data collection steps have already been performed and one wants to choose what data to collect next. After each step, the GRF model is updated using the cokriging equations \ref{eq:cokrig_mean} and \ref{eq:cokrig_cov} and we will denote by $\mathbb{P}_n$ the law of the field conditional on the data gathered up to (and including) stage $n$.

@@

Our goal is then to optimally select the next spatial location $\spatloc_{n+1}$ at which to observe the field (all of its component), where the meaning of optimal is yet to be defined. Note that here again, the concept of \textit{generalized location} allows one to consider settings where only one component of the field can be observed at each step and the startegy has to choose the spatial location as well as the component to observe (heterotopic case). In the same fashion, one could consider settings where several observations are allowed at each stage (batch case).

We will here restrict ourselves to the case where all components of the field are observed at one spatial location at each stage (isotopic case).

@@

At stage $s$, the robot must then choose design $d^{j,s}$ where measurements $\by_{d^{j,s}}$ should
be gathered. For the triangularized waypoint
graph all six possible
directions must be evaluated to minimize the expected uncertainty of random sets, and in particular the EIBV for our 
application. We hence focus on choosing designs for data modeled by
Eq. \eqref{eq:meas}, to conduct efficient sequential uncertainty
reduction according to the EIBV criterion. Compared with the
static design $d$ in Eq. \eqref{eq:eibv}, the optimal sequential
design not only considers the best current design grid point, but also
how the data gathered at this node could inform future sampling at
successive stages.

The mathematical expression for the optimal design involves a series
of intermixed maximizations over designs and integrals over data. In
practice, the optimal solution is intractable because of the enormous
growth over stages (see e.g. \cite{sucar2015probabilistic} and
\cite{powell2016perspectives}).  Instead, we outline heuristic
strategies.

@@

We now extend the myopic strategy to a look-ahead strategy. We consider two stages of
measurements, and the strategy is then optimal in the sense that it accounts consistently for the expectations and minimizations in these two stages, but still without including any planning beyond those two steps.
\midskip
Say we have collected data up to stage $n$. The principle of two-steps lookahead
strategies is to select the next observation location $\spatloc_{\stage + 1}$ as follows

  \begin{itemize}
      \item First, do as if the $\stage + 1$ observation had already been done and say
          we observed $y_{\stage + 1}$ at $\spatloc_{\stage + 1}$.
      \item If we were at this stage, then we would choose the next points by
          minimizing the expected IBV under the current ($\stage + 1$) conditional
          distribution
          \[
              \spatloc_{\stage + 2} = \argmin_{\spatloc \in \candidates}~ \EIBV_{\stage + 1}\left(\spatloc \mid \spatloc_{\stage + 1},
          y_{\spatloc + 1}\right)
          \]
      \item Now the above expression depends on the observed data $y_{\stage + 1}$,
          whose value we don't know beforehand. What we know is that
          at step $\stage$ it is distributed according to the conditional distribution
          $\mathbb{P}_{\stage}$. We can thus replace it by a random variable $Y$
          following the current conditional law and then average over it.
  \end{itemize}

  Let us now try to make the above more precise.
  \begin{criterion}[2-steps lookahead]
      The next observation location $\spatloc_{\stage + 1}$ is selected according to
      \begin{align}\label{critLA}
          \spatloc_{\stage + 1} = \argmin_{\spatloc \in \candidates}~ \mathbb{E}_{\stage}\left[\min_{\spatloc' \in
                  \candidates(\spatloc)} \EIBV_{\stage + 1}\left(\spatloc' \mid \spatloc,
      Y\right)\right]
      \end{align}
  where $Y$ is a random variable distributed according to the conditional
  law of $\gp_{\spatloc}$ at step $\stage$ (note it silently depends on the considered point
  $\spatloc$) and $\EIBV_{\stage +  1}(\spatloc'\mid \spatloc,Y)$ is the
  expected IBV with the expectation taken according to the conditional law $\mathbb{P}_{\stage}(\cdot\mid
  \spatloc, Y)$. Note that the dependence of the set of candidates on the current location has been made explicit for the second stage of measurements.
  \end{criterion}

\begin{remark}
In a practical setting, the first expectation might be computed by Monte
Carlo sampling of data $Y$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV provided b< Proposition \ref{propo2}, now with conditioning to the first stage data already going into the mean and covariances in Eq. \eqref{eq:krigSUR}. 
\end{remark}

>>>>

In a sequential setting, the point of view is that $n$ data collection steps have already been performed and one wants to choose what data to collect next. After each step, the GRF model is updated using the cokriging equations \ref{eq:cokrig_mean} and \ref{eq:cokrig_cov} and we will denote by $\mathbb{P}_n$ the law of the field conditional on the data gathered up to (and including) stage $n$. More generally, we will use subscripts to denote quantities conditional on the data available up to stage $\stage$. Hence, $\mu_{\stage}(\cdot)$ and $k_{\stage}(\cdot, \cdot)$ will denote conditional mean and covariance functions conditional on the first $\stage$ data collection stages.

@@

Our goal is then to optimally select what observations to do next, where the meaning of optimal is yet to be defined. Note that here again, the concept of \textit{generalized location} allows one to consider settings where only one component of the field can be observed at each step and the strategy has to choose the spatial location as well as the component to observe (heterotopic case), or settings where several observations are allowed at each stage (batch case).

We will here restrict ourselves to the case where only one spatial location can be chosen at each stage and all components of the field are observed there (isotopic case), leaving the general case for future work.

@@

\textcolor{red}{(C.T.) I suspect we need this paragraph so we cite people who have to be cited. How do we want to blend it with the rest? Personally I would get rid of this subsection.}
The mathematical expression for the optimal design involves a series
of intermixed maximizations over designs and integrals over data. In
practice, the optimal solution is intractable because of the enormous
growth over stages (see e.g. \cite{sucar2015probabilistic} and
\cite{powell2016perspectives}).  Instead, we outline heuristic
strategies.

@@

We now extend the myopic strategy to a look-ahead strategy. We consider two stages of
measurements, and the strategy is then optimal in the sense that it accounts consistently for the expectations and minimizations in these two stages, but still without including any planning beyond those two steps.

\medskip

Say we have collected data up to stage $n$. The principle of two-steps lookahead
strategies is to select the next observation location $\spatloc_{\stage + 1}$ that would yield the biggest exepcted decrease in IBV if we were to (optimally) add one more observation after the one at $\spatloc_{\stage + 1}$. In order to formalize this concept, we need to be able to talk about the EIBV in the future (after observation $\stage + 1$ has been made). The definition \ref{def:eibv} of $\currentEIBV$ includes a subscript to stress that expectations are taken with respect the conditional law at stage $n$. We will extend this notion by writing $\currentEIBV(\cdot; u, y)$ to denote EIBV where expectations are taken conditional on the data available at $n$ an on an additional observation $y$ at $u$.

  \begin{criterion}[2-steps lookahead]
      The next observation location $\spatloc_{\stage + 1}$ is selected according to
      \begin{align}\label{critLA}
          \spatloc_{\stage + 1} = \argmin_{\spatloc \in \candidates}~ \mathbb{E}_{\stage}\left[\min_{\spatloc' \in
                  \candidates(\spatloc)} \EIBV_{\stage}\left(\spatloc' ; \spatloc,
      Y\right)\right]
      \end{align}
  where $Y$ is a random variable distributed according to the conditional
  law of $\gp_{\spatloc}$ at step $\stage$ (note it silently depends on the considered point
  $\spatloc$). Here, the dependence of the set of candidates on the current location has been made explicit for the second stage of measurements.
  \end{criterion}

\begin{remark}
In a practical setting, the first expectation might be computed by Monte
Carlo sampling of data $Y$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV provided by Proposition \ref{propo2}, now with conditioning on the first stage data already going into the mean and covariances in Eqs. \eqref{eq:cokrig_mean} and \eqref{eq:cokrig_cov}.
\end{remark}

\begin{remark}
All our results have been formulated in a noiseless setting. That is, to us an observation $y$ at $\spatloc$ is a realization of $\gp_{\spatloc$. Generalization to a noisy setting where $y$ is a realization of $\gp_{\spatloc} + \epsilon$ with epsilon some noise process may be performed by including noise covariance matrices where necessary.
\end{remark}
